---
title: "PSPS Wave 1: Summary Statistics"
author: "Sarah Norman"
pagetitle: "PSPS Summary Statistics"

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    link-external-newwindow: true

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---

# Set Up

1. Needed Libraries
2. Loading Data Sets
3. Merging Data Sets

::: {.callout-note collapse="true"}

## Code Set Up

```{r}
#| label: Setup

## Libraries
library(haven)
library(tidyverse)
library(survey)
library(ggplot2)
library(showtext)
library(labelled)
library(scales)
library(here)

## Loading Data
household_survey <- read_dta(here("data/1_roster_edu_mig copy 2.dta"))
individual_survey <- read_dta(here("data/edu_policy_knowledge copy.dta"))
weights <- read_dta (here("data/design_weights copy.dta"))

```


```{r}
#| label: Cleaning
## Unify memid datatype
household_survey <- household_survey |>
  mutate(memid = as.character(memid))

individual_survey <- individual_survey |>
  mutate(memid = as.character(memid))

```


```{r}
#| label: Merging
## Merging
m_survey_1 <- household_survey |>
  left_join(
    individual_survey |>
      select(-fourp_status, -random_select, -nonrandom_select),
    by = c("province", "brgy_code", "hhid")
  )

m_survey_1 <- m_survey_1 |>
  left_join(
    weights,
    by = c("province", "brgy_code", "fourp_status")
  )

## did not match: 2163
sum(is.na(m_survey_1$norm_design_weight))
```

:::

## Data Manipulation and Weights

- Creation of catagorical`edu_level_group` variable: grouping `edu_completed` values into broader education levels (e.g., Elementary, Junior High, College+) while converting special missing codes (-999, -888, -666) to NA.
- Conversion of `edu_completed` to a factor variable with descriptive labels.
- Create binary indicator for college graduate or above
- Creation of `expected_school_level` variable: categorizing respondents into expected school levels based on their age (e.g., Elementary for ages 5-12, Junior High for ages 13-16).
- removal all rows where norm_design_weight is missing (NA).

::: {.callout-note collapse="true"}

# Code

```{r}
#| label: survey

# Created edu_level_group variable
m_survey_1 <- m_survey_1 |>
  mutate(
    edu_level_group = case_when(
      
      edu_completed == 0 ~ "No Schooling",
      
      edu_completed >= 1  & edu_completed <= 8  ~ "Elementary",
      
      edu_completed >= 9  & edu_completed <= 12 ~ "Junior High",
      
      edu_completed >= 13 & edu_completed <= 14 ~ "Senior High",
      
      edu_completed >= 15 & edu_completed <= 17 ~ "Vocational/Associate",
      
      edu_completed >= 18 & edu_completed <= 21 ~ "Some College",
      
      edu_completed >= 22 & edu_completed <= 23 ~ "College+",
      
      edu_completed %in% c(-999, -888, -666) ~ NA_character_,
      
      TRUE ~ NA_character_
    )
  )

m_survey_1 <- m_survey_1 |>
  mutate(
    edu_level_group = factor(
      edu_level_group,
      levels = c(
        "No Schooling",
        "Elementary",
        "Junior High",
        "Senior High",
        "Vocational/Associate",
        "Some College",
        "College+"
      )
    )
  )

## Conversion of `edu_completed` to a factor variable
m_survey_1$edu_completed <- haven::as_factor(
  m_survey_1$edu_completed
)

## Create binary indicator for college graduate or above
m_survey_1 <- m_survey_1 |>
  mutate(
    edu_completed = haven::as_factor(edu_completed),

    college_grad = case_when(
      is.na(edu_completed) ~ NA_real_,
      edu_completed %in% c("College graduate", "Education beyond college") ~ 1,
      TRUE ~ 0
    ),

    birth_year = 2023 - age_est,
    cohort10 = as.numeric(floor(birth_year / 10) * 10),

    k12_group = case_when(
      birth_year <= 1997 ~ "Before K-12",
      birth_year >= 1998 & birth_year <= 2000 ~ "Transition cohort",
      birth_year >= 2001 ~ "After K-12",
      TRUE ~ NA_character_
    ),

    completed_highschool = case_when(
      is.na(edu_completed) ~ NA_real_,
      edu_completed %in% c("Don't know", "Refuse to answer", "Other") ~ NA_real_,
      edu_completed %in% c(
        "10th Grade / HS Graduate / 4th year junior high school",
        "11th Grade",
        "12th Grade / SHS Graduate",
        "1st year of college",
        "2nd year of college",
        "3rd year of college",
        "4th year of college or higher",
        "College graduate",
        "Education beyond college",
        "1st Year Vocational training or associates degree",
        "2nd Year Vocational training or associates degree",
        "Vocational training or associates degree graduate"
      ) ~ 1,
      TRUE ~ 0
    )
  )

## Creation of `expected_school_level` variable
m_survey_1 <- m_survey_1 |>
  mutate(
    expected_school_level = case_when(
      age_est >= 5  & age_est <= 12 ~ "Elementary",
      age_est >= 13 & age_est <= 16 ~ "Junior High",
      age_est >= 17 & age_est <= 18 ~ "Senior High",
      age_est >= 19 & age_est <= 25 ~ "Post-Secondary",
      TRUE ~ NA_character_
    ),
    
    expected_school_level = factor(
      expected_school_level,
      levels = c(
        "Elementary",
        "Junior High",
        "Senior High",
        "Post-Secondary"
      )
    )
  )

## Removal all rows where norm_design_weight is missing (NA).

m_survey_clean <- m_survey_1 |>
  filter(!is.na(norm_design_weight))


```
:::

## Creating Weighted Object

```{r}
## Creating Survey Design Object

# Adjust variance estimates for strata containing only one PSU
options(survey.lonely.psu = "adjust")

# Define the complex survey design
# - Primary sampling units (PSUs): barangays and households
# - Stratification: province and 4Ps status
# - Sampling weights: normalized design weights
# - Finite population corrections (FPCs): barangay and household levels
design <- svydesign(
  id = ~brgy_code + hhid,
  strata = ~province + fourp_status,
  weights = ~norm_design_weight,
  fpc = ~fpc_brgy + fpc_hh,
  data = m_survey_clean,
  nest = TRUE
)
```


# Findings

1. Parents’ Education vs. Expectations for Children’s Education
2. Reasons for Not Applying for Tertiary Education Subsidies
3. College Graduation Rates by Birth Cohort
4. Reasons for Current Non-Enrollment
5. Reasons for Current Non-Enrollment by School Level

## Parents’ Education vs. Expectations for Children’s Education

::: {.callout-note collapse="true"}

# Code

```{r}
# Calculate the mean of ed_expect for each edu_level_group
edu_expect_completed <- svyby(
  ~ed_expect,
  ~edu_level_group,
  design,
  svymean,
  na.rm = TRUE
)

# Convert to a data frame for plotting
edu_expect_df <- as.data.frame(edu_expect_completed)

# Plotting
edu_expect_plot <- ggplot(
  edu_expect_df,
  aes(
    x = edu_level_group,
    y = ed_expect,
  ),
  color = "#397F94"
) +
  geom_point(size = 2) +
  geom_line(
    aes(group = 1),
    linewidth = 1,
    color = "#397F94"
  ) +
  geom_errorbar(
    aes(
      ymin = ed_expect - se,
      ymax = ed_expect + se
    ),
    width = 0.2
  ) +
  scale_y_continuous(
    breaks = c(18, 19, 20, 21, 22),
    labels = c(
      "1st Year\nCollege",
      "2nd Year\nCollege",
      "3rd Year\nCollege",
      "4th Year+\nCollege",
      "College\nGraduate"
    )
  ) +
  labs(
    x = "Parent's Highest Education Completed",
    y = "Average Parent Expectation \nfor Child's Educational Attainment",
    title = "Parents With Higher Education \nExpect Higher Education for Their Children",
    caption = "Weighted PSPS Survey Data"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

```

:::

```{r}
#| echo: false
edu_expect_plot
```

## Reasons for Not Applying for Tertiary Education Subsidies

::: {.callout-note collapse="true"}

# Code

```{r}
# Weighted proportions
subsidy_weighted <- svymean(
  ~avail_1_b_complex_process +
    avail_1_b_expect_discontinue +
    avail_1_b_has_funds +
    avail_1_b_inelig_finNeed +
    avail_1_b_low_priority +
    avail_1_b_noInterest_HE +
    avail_1_b_noInterest_elig_insti +
    avail_1_b_oth_needier +
    avail_1_b_other_support,
  design = design,
  na.rm = TRUE
)

# Convert to dataframe
subsidy_df <- tibble(
  reason = names(coef(subsidy_weighted)),
  percent = as.numeric(coef(subsidy_weighted)) * 100,
  se = as.numeric(SE(subsidy_weighted)) * 100
) |>
  mutate(
    reason = case_when(
      reason == "avail_1_b_complex_process" ~ "Process too complicated",
      reason == "avail_1_b_expect_discontinue" ~ "Expect program to discontinue",
      reason == "avail_1_b_has_funds" ~ "Has sufficient funds",
      reason == "avail_1_b_inelig_finNeed" ~ "Financial need criteria",
      reason == "avail_1_b_low_priority" ~ "Unlikely to be selected",
      reason == "avail_1_b_noInterest_HE" ~ "Student uninterested",
      reason == "avail_1_b_noInterest_elig_insti" ~ "Won't attend eligible school",
      reason == "avail_1_b_oth_needier" ~ "Other families more deserving",
      reason == "avail_1_b_other_support" ~ "Seeking other support",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(reason)) |>
  arrange(desc(percent))

# Plot
subsidy_plot <- ggplot(
  subsidy_df,
  aes(
    x = reorder(reason, percent),
    y = percent
  )
) +
  geom_col(fill = "#397F94") +
  geom_text(
    aes(label = paste0(round(percent, 1), "%")),
    hjust = -0.2,
    size = 4
  ) +
  coord_flip() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = label_percent(scale = 1)
  ) +
  labs(
    x = "",
    y = "Weighted Percentage of Respondents",
    title = "Reasons for Not Taking Up the Tertiary Education Subsidy",
    caption = "Weighted PSPS Survey Data"
  ) +
  theme_minimal(base_size = 13)
```

:::

```{r}
#| echo: false
subsidy_plot 
```


## College Graduation Rates by Birth Cohort

::: {.callout-note collapse="true"}

## Code

```{r}

college_by_cohort <- svyby(
  ~college_grad,
  ~cohort10,
  subset(design, cohort10 >= 1930 & cohort10 <= 2000),
  svymean,
  na.rm = TRUE
)

## Graphing results
# number of college graduates is increasing overtime
college_grads_graph <- ggplot(college_by_cohort, aes(
    x = factor(cohort10), 
    y = college_grad * 100, 
    fill = college_grad * 100)) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = (college_grad - se) * 100,
      ymax = (college_grad + se) * 100
    ),
    width = 0.2
  ) +
  scale_fill_gradientn(
    colours = c("#D6EAF0", "#7FB3C2", "#397F94"),
    values = c(0, 0.1, 1)
  ) +
  labs(
    x = "Birth Cohort",
    y = "Percent of College Graduate or Beyond",
    title = "Percent College Graduate or Beyond by Birth Cohort",
    caption = "Data: PSPS 2023 Household Survey"
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 18,
      hjust = 0.5),
    plot.margin = margin(
        t = 30,  # top
        r = 30,  # right
        b = 30,  # bottom
        l = 30   # left
      ), 
    legend.position = "none"
    )

```

:::

```{r}
#| echo: false
college_grads_graph
```

## Reasons for Current Non-Enrollment

::: {.callout-note collapse="true"}

## Code

```{r}

# Edu_reasons1 analysis ----

## Means
edu_reasons <- svymean(
  ~edu_reasons1_NoMoney +
    edu_reasons1_Job +
    edu_reasons1_Marriage +
    edu_reasons1_Disability +
    edu_reasons1_Sick +
    edu_reasons1_HomeObligation +
    edu_reasons1_NoSchool_Teacher +
    edu_reasons1_NoTime +
    edu_reasons1_ParentsDeath +
    edu_reasons1_ParentsSep +
    edu_reasons1_Pregnant,
  
  subset(design, age_est > 5 & age_est < 26),
  na.rm = TRUE
)

edu_reasons

## Creating dataframe
reasons_df <- data.frame(
  reason = names(coef(edu_reasons)),
  percent = coef(edu_reasons) * 100,
  se = SE(edu_reasons) * 100
)

## Clean labels
reasons_df$reason <- c(
  "No Money",
  "Need to Work",
  "Marriage",
  "Disability",
  "Sick",
  "Home Obligations",
  "No School / Teacher",
  "No Time",
  "Parents Died",
  "Parents Separated",
  "Pregnant"
)

## Graphing results
edu_reasons_graph <- ggplot(
  reasons_df,
  aes(
    x = reorder(reason, percent),
    y = percent,
    fill = percent
  )
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = percent - se,
      ymax = percent + se
    ),
    width = 0.2
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.12)),
    breaks = seq(0, 30, by = 5),
    limits = c(0, 30)
    ) +
  scale_fill_gradient(
    low = "#D6EAF0",
    high = "#397F94"
  ) +
  labs(
    x = "",
    y = "Percent",
    title = "Reasons for Not Currently Enrolled in School",
    caption = "Ages 6–25 | Weighted PSPS Survey Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    legend.position = "none",
    plot.margin = margin(
      t = 30,  # top
      r = 50,  # right
      b = 30,  # bottom
      l = 5   # left
    ),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  geom_text(
    data = reasons_df |> slice_max(percent, n = 4),
    aes(label = paste0(round(percent, 1), "%")),
    hjust = -1,
    fontface = "bold"
  )

```

:::

```{r}
#| echo: false
edu_reasons_graph
```


## Reasons for Current Non-Enrollment by School Level

::: {.callout-note collapse="true"}

## Code

```{r}

## Means by school level
edu_reasons_by_level <- svyby(
  ~edu_reasons1_NoMoney +
    edu_reasons1_Job +
    edu_reasons1_Marriage +
    edu_reasons1_Disability +
    edu_reasons1_Sick +
    edu_reasons1_HomeObligation +
    edu_reasons1_NoSchool_Teacher +
    edu_reasons1_NoTime +
    edu_reasons1_ParentsDeath +
    edu_reasons1_ParentsSep +
    edu_reasons1_Pregnant,
  ~expected_school_level,
  subset(design, age_est >= 5 & age_est <= 25),
  svymean,
  na.rm = TRUE
)

edu_reasons_by_level

## Creating dataframe
reasons_df_level <- edu_reasons_by_level |>
  as.data.frame() |>
  pivot_longer(
    cols = starts_with("edu_reasons1_"),
    names_to = "reason",
    values_to = "percent"
  ) |>
  pivot_longer(
    cols = starts_with("se."),
    names_to = "se_reason",
    values_to = "se"
  ) |>
  mutate(
    reason = str_remove(reason, "edu_reasons1_"),
    se_reason = str_remove(se_reason, "se.edu_reasons1_")
  ) |>
  filter(reason == se_reason) |>
  mutate(
    percent = percent * 100,
    se = se * 100
  )

## Graphing by class

edu_reasons_graph_year <- reasons_df_level |>
  filter(expected_school_level != "Post-Secondary") |>
  ggplot(
    aes(
      x = reorder(reason, percent),
      y = percent,
      fill = percent
      )
    )+
  geom_col() +
  geom_errorbar(
    aes(
      ymin = percent - se,
      ymax = percent + se
    ),
    width = 0.2
  ) +
  coord_flip() +
  facet_wrap(~expected_school_level) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.12)),
    breaks = seq(0, 100, by = 10)
  ) +
  scale_fill_gradient(
    low = "#D6EAF0",
    high = "#397F94"
  ) +
  labs(
    x = "",
    y = "Percent",
    title = "Reasons for Not Currently Enrolled \nin School by School Level",
    caption = "Ages 6–25 | Weighted PSPS Survey Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines")
  ) +
  coord_flip(clip = "off") +
  geom_text(
    data = reasons_df_level |>
      filter(expected_school_level != "Post-Secondary") |>
      group_by(expected_school_level) |>
      slice_max(percent, n = 4) |>
      ungroup(),
    
    aes(label = paste0(round(percent, 1), "%")),
    hjust = -0.85,
    fontface = "bold",
    size = 3
  )
```

:::

